Cosmological Hydrodynamic Simulations of Preferential Accretion in the SMBH of Milky Way Size Galaxies

### 2016-05-20 ms.tex
Jillian Edits - May 16th

\title{A Particular Appetite: Cosmological Hydrodynamic Simulations of Preferential Accretion in the Supermassive Black Holes of Milky Way Size Galaxies}

@@

[IAU Abstract as Placeholder] Using cosmological hydrodynamic simulations of Milky Way-type Galaxies, we explore the effect of accreted gas as feeding mechanisms for supermassive black holes. By examining two of these galaxies with differing merger histories, one characterized by several major mergers and the other with a quiescent history, we can examine the importance of merger history on black hole accretion. This study is an extension of Bellovary et. al. 2013, which did a similar study analyzing the accretion of high mass, high redshift galaxies and their central black holes. Bellovary found that the gas accreted by the central black holes was proportional to that accreted by the host galaxy. Contrary to the previous study's results, we've found that while a galaxy with a quiescent history will still have a black hole mirroring the accretion of its host, a galaxy with an active merger history has a central black hole that is preferentially fueled by gas accreted through mergers. We look to the the angular momentum of the accreted gas in these Milky Way analogs to develop a clearer picture of the mechanisms best fueling their central SMBHs.

@@

The Amiga Halo Finder identifies all of the halos in the simulations by identifying the virial radii of the halos based on an overdensity using the criterion for a flat universe. \citep{Knebe2001,Knollmann2009,} We select the primary halo by determining the most massive galaxy at z=0 and locating the central SMBH. For our Gasoline simulations, both the primary halo in h258 and h277 had final masses on the order of $10^7 M_{\odot}$ and a formation redshift of z~4.

@@

The galaxy, h258, is characterized by an active merger history including a major merger at z=1. The merger can be seen affecting the cumulative black hole mass (Figure \ref{h258allmassgas}) as well as the gas fractions within the both the SMBH and galaxy (Figure \ref{h258numfrac}). 
Figure \ref{h258allmassgas} shows h258's central SMBH's cumulative mass as a function of time (lower axis) and redshift (upper axis). The black dashed line indicates the total cumulative BH mass (including both mass from gas and BH mergers), while the black solid line indicates only the total gas mass accreted by the BH. The blue dot-dashed line indicates the gas mass accreted via unshocked gas. The green solid line indicates the gas mass accreted through mergers. The red dashed line indicates gas mass that was shocked upon entry into the halo. Though this figure shows that the primary source of mass gain at high redshifts is through mergers with other black holes, there is a significant increase in the amount of gas adding to the mass of the SMBH starting at z $\sim$ 1, where we know a major merger takes place. \textbf{[Reference merger tree here once you've added it.]} Similarly, Figure \ref{h258numfrac} shows that unshocked gas (blue) makes up the majority of gas entering the galaxy (solid lines) at early times; nevertheless, around the time of the merger (z $\sim$ 1) a marked change occurs in the gas fractions within the black hole (dashed lines) where clumpy gas (green) becomes the main component of total gas.

Figure \ref{h258piefrac} depicts the fractions of total gas in the galaxy (a) and the SMBH (b) at z=0, again differentiated by the gas particles' origins. Blue, green, and red distinguish cold, clumpy, and shocked, respectively. Yellow indicates gas that existed within the main halo upon formation. The galaxy has a mass nearly half comprised of gas entering the galaxy through unshocked, smooth accretion (48 \%), with 39 \% of the gas entering through mergers. The smallest fractions of the total gas are comprised of shocked gas (9 \%) and ``early''gas which existed at formation makes up the final, smallest fraction (4 \%). Unlike the galaxy, nearly 65 \% of the gas accreted by the central SMBH was gas accreted via mergers, while only a quarter (26 \%) is comprised of unshocked, smoothly accreted gas. Shocked gas makes up the last 9 \% of total gas entering the SMBH, along with a negligible ($<$ 1\%) amount of ``early'' gas. \textbf{So Figure \ref{h258piefrac} shows explicitly that the SMBH more readily accretes gas gained through mergers.} This result is \textit{contrary} to the previous results of \cite{Bellovary2013}! This result is surprising, but important, as it allows us to consider that a different physical mechanism characterizes the gas entering this MW-type galaxy than those affecting the gas accretion of the high mass galaxies examined by \cite{Bellovary2013}.

We examine the angular momentum of the gas at the moment it enters the galaxy for clues to the understand the ready accretion of merger accreted gas. (Figure \ref{h258angmom}) A cumulative distribution of the angular momentum of entering gas, differentiated based on method of entrance into the galaxy, shows that gas entering the SMBH (dashed lines) has an overall lower angular momentum than gas entering galaxy (solid lines). This statement is true whether the gas enters via unshocked smooth accretion, becomes shocked upon entry, or enters through a galaxy merger (blue, red, green, respectively). Therefore, we can infer from this plot that lower angular momentum gas more readily flows into the central region of the host galaxy where it can be accreted by the black hole. In contrast, \cite{Bellovary2013} found a \textit{slight} preference for gas of a lower angular momentum to enter the SMBH. The question still remains as to why clumpy gas in particular appears most prominently in the SMBH at z = 0. The answer to this conundrum may lie in the major merger that characterizes this galaxy, so we look specifically at the gas accreted during the merger.

@@

@@

This result can be more clearly seen in Figure \ref{h277piefrac}. (Colors as in \ref{h258piefrac}.) Similar to Bellovary et. al. 2013 study, which found SMBH accretion in high mass, high redshift galaxies mirror their hosts in terms of accreted gas, h277's SMBH accretes a similar fraction of each type of gas, comparable to its host galaxy. Smoothly accreted, unshocked gas represents about half of the total gas in both the SMBH and the galaxy. Clumpy gas also maintains about 30\% of both, while an increase in shocked gas occurs, from 13 \% in the galaxy to 22 \% in the SMBH. %; however, we argue this may be due to supernova contamination in the simulation. 
So by present day, h277 found no strong preference for the SMBH to prefer any one kind of gas in its accretion. Like the SMBHs in the study of \cite{Bellovary2013}, nearly the same fractions of gas entering the halo would then be accreted onto the SMBH (with the exception of shocked gas). Therefore, this may imply that without some physical driver, like the major merger of h258, clumpy gas is unlikely to dominate the gas accreted into the SMBH.

As in the previous case, we examined the angular momentum of all the gas particles accreted by the SMBH and host galaxy at the moment of entry. (Figure \ref{h277angmom}; Linestyles as in \ref{h258angmom}) Like before, it is immediately apparent that the merger gas accreted by both the black hole and its host has a lower angular momentum overall than the smoothly accreted gas; however, unlike we found in h258, the h277 galaxy accretes gas with lower angular momentum than that of its SMBH. This means that even though some of the gas in the host halo had lower angular momentum, that gas didn't find its way into the central region to accrete onto the BH. This may again imply that some physical phenomena is required to drive those gases towards the center.

@@

\begin{figure}
\centerline{\resizebox{0.55\hsize}{!}{\includegraphics[angle=0]{PIE_h258galgasfrac_noexplode_realcolors}}}
\centerline{\resizebox{0.55\hsize}{!}{\includegraphics[angle=0]{PIE_h258bhgasfrac_noexplode_realcolors}}}
\caption[]{Gas fractions of the gas particles accreted in h258 by the main halo (above) and the SMBH (below), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligible ($<$ 1 $\%$) within the SMBH.}
\label{h258piefrac} 
\end{figure}

@@

\begin{figure}
\centerline{\resizebox{0.55\hsize}{!}{\includegraphics[angle=0]{PIE_h277galgasfrac_noexplode_realcolors}}}
\centerline{\resizebox{0.55\hsize}{!}{\includegraphics[angle=0]{PIE_h277bhgasfrac_noexplode_realcolors}}}
\caption[]{Gas fractions of the gas particles accreted in h258 by the main halo (above) and the SMBH (below), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligable (!! 1 \%) within the SMBH.}
\label{h277piefrac} 
\end{figure}

@@

>>>>

\title{Cosmological Hydrodynamic Simulations of Preferential Accretion in the SMBH of Milky Way Size Galaxies}
%\title{A Particular Appetite: Cosmological Hydrodynamic Simulations of Preferential Accretion in the Supermassive Black Holes of Milky Way Size Galaxies}

@@

Using cosmological hydrodynamic simulations of Milky Way-type Galaxies, we explore the effect of accreted gas as feeding mechanisms for supermassive black holes. We examine two of these galaxies with differing merger histories using the Gasoline N-body+SPH code. One of there galaxies is characterized by several major mergers and the other with a quiescent history. We attempt to identify the importance of merger history on black hole accretion. Additionally, we consider a second, higher resolution version of the Milky Way-type galaxy with an active merger history using the Charm N-body GrAvity solver, ChaNGa.  This study is an extension of Bellovary et. al. 2013, which analyzed the accretion of high mass, high redshift galaxies and their central black holes. Bellovary found that the gas accreted by the central black holes was proportional to that accreted by the host galaxy. Contrary to the previous study's results, we've found that while a galaxy with a quiescent history will still have a SMBH whose accretion mirrors that of its host, a galaxy with an active merger history has a central black hole that is preferentially fueled by gas accreted through mergers. Through an investigation of the angular momentum of the gas entering these hosts and their SMBHs, we determine that merger gas enters the galaxy with lower angular momentum compared to smooth accretion, accounting for the preferential fueling witnessed in the SMBH of the galaxy with an active merger history. Our results imply that galaxy mergers play an important role in the feeding of the SMBHs in the class of Milky Way-type galaxies.

@@

The Amiga Halo Finder identifies all of the halos in the simulations by identifying the virial radii of the halos based on an overdensity using the criterion for a flat universe. \citep{Knebe2001,Knollmann2009,Gill2004} We select the primary halo by determining the most massive galaxy at z=0 and locating the central SMBH. For our Gasoline simulations, both the primary halo in h258 and h277 had final masses on the order of $10^{12} M_{\odot}$ and a formation redshift of z~4.

@@

The galaxy, h258, is characterized by an active merger history including a major merger at z=1. The merger can be seen affecting the cumulative black hole mass (Figure \ref{h258allmassgas}) as well as the gas fractions within the both the SMBH and galaxy (Figure \ref{h258numfrac}). 
Figure \ref{h258allmassgas} shows h258's central SMBH's cumulative mass as a function of time (lower axis) and redshift (upper axis). The black dashed line indicates the total cumulative BH mass (including both mass from gas and BH mergers), while the black solid line indicates only the total gas mass accreted by the BH. The blue dot-dashed line indicates the gas mass accreted via unshocked gas. The green solid line indicates the gas mass accreted through mergers. The red dashed line indicates gas mass that was shocked upon entry into the halo. Though this figure shows that the primary source of mass gain at high redshifts is through mergers with other black holes, there is a significant increase in the amount of gas adding to the mass of the SMBH starting at z $\sim$ 1, where we know a major merger takes place. \textbf{[Reference merger tree here once you've added it.]} Figure \ref{h258numfrac} shows the fractions of unshocked and clumpy gas compared to all accreted gas as a function of redshift. It confirms that unshocked gas (blue) makes up the majority of gas entering the galaxy (solid lines) at early times; nevertheless, around the time of the merger (z $\sim$ 1) a marked change occurs in the gas fractions within the black hole (dashed lines) where clumpy gas (green) becomes the main component of total gas.

Figure \ref{h258stackfrac} depicts the fractions of total gas in the galaxy (a) and the SMBH (b) at z=0, again differentiated by the gas particles' origins. Blue, green, and red distinguish cold, clumpy, and shocked, respectively. Yellow indicates gas that existed within the main halo upon formation. The galaxy has a mass nearly half comprised of gas entering the galaxy through unshocked, smooth accretion (48 \%), with 39 \% of the gas entering through mergers. The smallest fractions of the total gas are comprised of shocked gas (9 \%) and ``early''gas which existed at formation makes up the final, smallest fraction (4 \%). Unlike the galaxy, nearly 65 \% of the gas accreted by the central SMBH was gas accreted via mergers, while only a quarter (26 \%) is comprised of unshocked, smoothly accreted gas. Shocked gas makes up the last 9 \% of total gas entering the SMBH, along with a negligible ($<$ 1\%) amount of ``early'' gas. \textbf{So Figure \ref{h258stackfrac} shows explicitly that the SMBH more readily accretes gas gained through mergers.} This result is \textit{contrary} to the previous results of \cite{Bellovary2013} which found that the fractions of gas comprising the SMBH's total gas mass and that in its host were nearly the same. This result is surprising, but important, as it allows us to consider that a different physical mechanism characterizes the gas entering this MW-type galaxy than those affecting the gas accretion of the high mass galaxies examined by \cite{Bellovary2013}.

We examine the angular momentum of the gas at the moment it enters the galaxy for clues to the understand the ready accretion of merger accreted gas. (Figure \ref{h258angmom}) A cumulative distribution of the angular momentum of entering gas, differentiated based on method of entrance into the galaxy, shows that gas entering the SMBH (dashed lines) has an overall lower angular momentum than gas entering galaxy (solid lines). This statement is true whether the gas enters via unshocked smooth accretion, becomes shocked upon entry, or enters through a galaxy merger (blue, red, green, respectively). Therefore, we can infer from this plot that lower angular momentum gas more readily flows into the central region of the host galaxy where it can be accreted by the black hole. In contrast, \cite{Bellovary2013} found a \textit{slight} preference for gas of a lower angular momentum to enter the SMBH. The question remains, however, as to why unshocked gas contains some of the lowest angular momentum gas while clumpy gas nevertheless appears most prominently in the SMBH at z = 0. The answer to this conundrum may lie in the major merger that characterizes this galaxy, so we look specifically at the gas accreted during the merger.

@@

\textit{NOTE: Need to add a figure that clearly supports this argument. Possibly a series of angular momentum plots at the times of mergers. Or comparisons between cumulative distributions at time with and without mergers.}

@@

This result can be more clearly seen in Figure \ref{h277stackfrac}. (Colors as in \ref{h258stackfrac}.) Similar to Bellovary et. al. 2013 study, which found SMBH accretion in high mass, high redshift galaxies mirror their hosts in terms of accreted gas, h277's SMBH accretes a similar fraction of each type of gas, comparable to its host galaxy. Smoothly accreted, unshocked gas represents about half of the total gas in both the SMBH and the galaxy. Clumpy gas also maintains about 30\% of both, while an increase in shocked gas occurs, from 13 \% in the galaxy to 22 \% in the SMBH. %; however, we argue this may be due to supernova contamination in the simulation. 
So by present day, h277 found no strong preference for the SMBH to prefer any one kind of gas in its accretion. Like the SMBHs in the study of \cite{Bellovary2013}, nearly the same fractions of gas entering the halo would then be accreted onto the SMBH (with the exception of shocked gas). Therefore, this may imply that without some physical driver, like the major merger of h258, clumpy gas is unlikely to dominate the gas accreted into the SMBH.

As in the previous case, we examined the angular momentum of all the gas particles accreted by the SMBH and host galaxy at the moment of entry. (Figure \ref{h277angmom}; Linestyles as in \ref{h258angmom}) Like before, it is immediately apparent that the clumpy gas accreted by both the black hole and its host has a lower angular momentum overall than the smoothly accreted gas; however, unlike we found in h258, the h277 galaxy accretes gas with lower angular momentum than that of its SMBH. This appears to imply that some of the gas with the lowest angular momentum is remaining in the host galaxy rather than accreting onto the SMBH.

% IS THIS WHAT IT MEANS? This means that even though some of the gas in the host halo had lower angular momentum, that gas didn't find its way into the central region to accrete onto the BH. This may again imply that some physical phenomena is required to drive those gases towards the center.

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{h258_stackbarfractions}}}
\caption[]{Gas fractions of the gas particles accreted in h258 by the main halo (left) and the SMBH (right), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligible ($<$ 1 $\%$) within the SMBH.}
\label{h258stackfrac} 
\end{figure}

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{h277_stackbarfractions}}}
\caption[]{Gas fractions of the gas particles accreted in h258 by the main halo (left) and the SMBH (right), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligable ($<$ 1 \%) within the SMBH.}
\label{h277stackfrac} 
\end{figure}

@@

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_massgastime}}}
\caption[]{The central BH’s cumulative mass as a function of time and redshift. The black dashed line indicates the total cumulative BH mass. The black solid line indicates the total gas mass. The blue dot-dashed line indicates the gas mass accreted via unshocked gas. The green solid line indicates the gas mass accreted through mergers. The red dashed line indicates gas mass that was shocked upon entry into the halo.}
\label{hrh258allmassgas} 
\end{figure}

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_am_cumu_4096}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto the high resolution ChaNGa h258.  Gas particles accreted onto the main halo (solid lines) and central black hole (dashed lines). The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{hrh258angmom} 
\end{figure}

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_am_tsnew_1584}}}
\caption[]{ Cumulative distribution of angular momentum of the gas particles accreted onto the high resolution ChaNGA h258 galaxy at the time of the major merger (z ~ 1). There are about 450,000 gas particles accreted at this timestep, which gas particles accreted onto the main halo and central black hole distinguished by solid and dashed lines. The green, blue, and red lines indicate clumpy, unshocked, and shocked gas, respectively.}
\label{hrh258angmom_merger} 
\end{figure}

\begin{figure}
\centerline{\resizebox{0.75\hsize}{!}{\includegraphics[angle=0]{hrh258_stackbarfractions}}}
\caption[]{Gas fractions of the gas particles accreted by the high resolution ChaNGA h258 by the main halo (left) and the SMBH (right), distinguished by type. Blue, green, and red distinguish gas gained through unshocked gas, gained through mergers, and gas shocked upon entry, respectively. Yellow indicates gas that existed within the main halo upon formation; this ``early'' gas is negligable ($<$ 1 \%) within the SMBH.}
\label{h277stackfrac} 
\end{figure}
